---
title: "My best nature shots"
description: |
   
output:
  distill::distill_article
base_url: https://souzayuri.github.io/
---
```{r echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE
)
```
<head>
<link type="text/css" rel="stylesheet" href="https://cdnjs.cloudflare.com/ajax/libs/lightgallery/1.6.0/css/lightgallery.min.css" />
<script src="https://cdnjs.cloudflare.com/ajax/libs/lightgallery-js/1.4.1-beta.0/js/lightgallery.min.js"></script>
<!-- lightgallery plugins -->
<script src="https://cdnjs.cloudflare.com/ajax/libs/lg-fullscreen/1.2.1/lg-fullscreen.min.js"></script>
<script src="https://cdnjs.cloudflare.com/ajax/libs/lg-thumbnail/1.2.1/lg-thumbnail.min.js"></script>
<script src="https://cdnjs.cloudflare.com/ajax/libs/lg-zoom/1.3.0/lg-zoom.min.js"></script>

<!-- Add Font Awesome for the magnifier icon -->
<link rel="stylesheet" href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.15.4/css/all.min.css" />

<!-- Add custom CSS for magnifier -->
<style>
.img-container {
  position: relative;
  display: inline-block;
}

.magnifier-icon {
  position: absolute;
  bottom: 5px;
  right: 5px;
  background-color: rgba(0, 0, 0, 0.5);
  color: white;
  border-radius: 50%;
  width: 30px;
  height: 30px;
  display: flex;
  align-items: center;
  justify-content: center;
  opacity: 0;
  transition: opacity 0.3s;
  cursor: pointer;
}

.img-container:hover .magnifier-icon {
  opacity: 1;
}

.lg-zoomed .lg-image {
  cursor: move !important;
}
</style>
</head>

```{css}
#lightgallery > a > img:hover {
   transform: scale(1.15, 1.15);
   transition: 0.4s ease-in-out;
   cursor: pointer;
}
```
```{r include = FALSE}
# Load the functions we have created
if (!require("here")) install.packages("here")
source(here::here("R/functions_nature.R"))
```
```{r}
# Create layout
withTags(
  div(
    class = "row",
    id = "lightgallery",
    tagList(
      make_gallery_layout()
    )
  )
)
```
<script type="text/javascript">
    // Initialize lightGallery with zoom options
    lightGallery(document.getElementById('lightgallery'), {
        zoom: true,
        scale: 1.2,
        actualSize: false,
        download: false,
        addClass: 'lg-custom-zoom'
    });
    
    // Add magnifier icons to all gallery items after page loads
    document.addEventListener('DOMContentLoaded', function() {
        const galleryItems = document.querySelectorAll('#lightgallery a');
        
        galleryItems.forEach(function(item) {
            // Convert simple <a> with <img> to container with icon
            const imgElement = item.querySelector('img');
            const imgContainer = document.createElement('div');
            imgContainer.className = 'img-container';
            
            // Create magnifier icon
            const magnifier = document.createElement('div');
            magnifier.className = 'magnifier-icon';
            magnifier.innerHTML = '<i class="fas fa-search-plus"></i>';
            
            // Create a clone of the image to place in our container
            const imgClone = imgElement.cloneNode(true);
            
            // Structure: <div class="img-container"><img /><div class="magnifier-icon">...</div></div>
            imgContainer.appendChild(imgClone);
            imgContainer.appendChild(magnifier);
            
            // Replace the original image with our container
            imgElement.parentNode.insertBefore(imgContainer, imgElement);
            imgElement.remove();
            
            // Add click event to the magnifier icon
            magnifier.addEventListener('click', function(e) {
                e.preventDefault();
                e.stopPropagation();
                
                // Get the image index for lightGallery
                const index = Array.from(galleryItems).indexOf(item);
                
                // Open lightGallery at this index
                window.lgData[document.getElementById('lightgallery').getAttribute('lg-uid')].goToSlide(index);
                
                // Trigger zoom after a short delay to ensure slide is loaded
                setTimeout(function() {
                    // Get the current slide's zoom toggle button and click it
                    const zoomToggle = document.querySelector('.lg-zoom-in');
                    if (zoomToggle) {
                        zoomToggle.click();
                    }
                }, 300);
            });
        });
    });
</script>
